Fire danger: the skill provided by the ECMWF Integrated Forecasting System

### 2019-11-11 article/article.tex
Added figures1-3

---
title: "Fire danger: the skill provided by the ECMWF Integrated Forecasting System"
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
author:
  - given_name: Francesca
    surname: Di Giuseppe
    affiliation: 1
    email: F.DiGiuseppe@ecmwf.int
    corresponding: true
  - given_name: Claudia
    surname: Vitolo
    affiliation: 1
  - given_name: Blazej
    surname: Krzeminski
    affiliation: 1
  - given_name: Jesus
    surname: San-Miguel
    affiliation: 2
  - given_name: Florian
    surname: Pappenberger
    affiliation: 1
affiliation:
  - code: 1
    address: European Centre for Medium-range Weather Forecasts, Reading, United Kingdom
  - code: 2
    address: European Commission, Joint Research Centre, Ispra, Italy
abstract: "In the framework of the EU Copernicus program the European Centre for Medium-range Weather Forecast (ECMWF) on behalf of the Joint Research Centre (JRC) is forecasting daily fire weather indices using its medium range ensemble prediction system. The use of weather forecast in place of local observations can extend early warnings up to 1-2 weeks allowing for greater proactive coordination of resource-sharing and mobilization within and across countries. In addition, the use of an ensemble system allows to perform a probabilistic assessment of the forecast uncertainties which can boost confidence in the decision process during emergency situations. Using one year of pre-operational service in 2017 here we assess the capability of the system globally and analyze in detail three major events in Chile, Portugal and California. We also present examples on how fire forecast products could be tailored to provide uncertainties of fire weather conditions and information in a probabilistic fashion."
bibliography: references.bib
running:
  title: Fire danger forecasting
  author: Di Giuseppe et al.
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  The authors declare no competing interests.
availability:
  code: "The code to generate the results of this work is available as vignette of the caliver R package: Verification of fire danger classes."
authorcontribution: |
  FDG wrote the ...
acknowledgements: |
  This work was founded by the EU Project ANYWHERE (Contract 700099) and the Global Fire Contract 389730 between the Joint Research Centre and ECMWF.
output:
  rticles::copernicus_article: default
  bookdown::pdf_book:
    base_format: rticles::copernicus_article
---

@@

Systems such as the FWI detect dangerous weather conditions conducive of uncontrollable fires rather than modelling the probability of ignition and fire behaviours. The FWI (developed in Canada) is specifically calibrated to describe the fire behaviour in a jack pine stand (*Pinus banksiana*) typical of the Canadian forests. However, its simplicity of implementation has made it a popular choice in many countries and it has shown to perform reasonably well in ecosystems very dissimilar to the boreal forest \citep{digiuseppe:16,degroot:07}. The FWI calculation only relies on weather forcings and no information on the actual vegetation status is taken into account. When weather forecasts are used in place of observations, uncertainties can be introduced. Sources of uncertainty can be: (i) the limited predictability of atmospheric dynamics which is strongly affected by the initial state and (ii) the misrepresentation of physical processes. In the former case, errors are randomly distributed around the true state \citep{orrell:01}; in the latter, errors produce systematic deviations from the true state. In both cases, errors in the weather forecast may be amplified or damped by nonlinear transformations in the fire weather model \citep{erickson:18}. Thus, for example, a dry bias in the model in a certain region will lead to the persistent prediction of higher fire danger values compared to what would be calculated using local observations.

@@

\subsection{FWI calculation}
\subsubsection{General concept}
The Fire Weather Index system is composed of six variables describing fuel moisture and fire behavior, as influenced by weather \citep{vanwagner:87}. Three fuel moisture codes provide the moisture content of dead woody debris of different diameter classes laying on three fuel beds. The dead fine fuels is described by the Fine Fuel Moisture Code (FFMC) which has a fast response time to weather modification (hourly time scales). The surface organic matter of moderate density, such as the fermentation layer of forest soils \citep{hood:10} is calculated using the Duff Moisture Code (DMC). DMC has a slower response to the weather forcing, conventionally set to 12 days. Finally the deep, compact soil organic layers, such as the humus layer of forest soils \citet{hood:10} is expressed with a Drought Code (DC). This is the slower evolving component of the FWI codes and has a lag-time set to 52 days in the original FWI formulation. The FWI system also provides fire behavior indexes, in terms of rate of fire spread (Initial Spread Index, ISI), fuel available for combustion (Buildup Index, BUI), and head fire intensity (Fire Weather Index, FWI). The FWI component combines the ISI and BUI and can be considered as the most general indicator of fire danger.

@@

In addition, from ERA5 we derive a climatological benchmark simulation (called CLIM hereafter), considered as a "mean year" for the FWI using the period 1980-2018. At pixel level, every day the median FWI is calculated over the period spanning 4 days before/after a given date. While CLIM has no expected predictive skill for a specific year, it retains information on the yearly climatological variation of FWI. CLIM is what could be used in the absence of either a monitoring (based on observations) or a forecast system (based on weather forecast). It should therefore score better or equal of the forecast on time ranges beyond the limits of predictability. CLIM is used in this study as a benchmark to rank the expected improvements provided by a forecasting system

@@

While national inventories of wildfire activities exist in many countries, they can be heterogeneous and lack the temporal span desirable for the validation of a fire danger system at the global scale. Satellite observations can supply a valid alternative especially as they cover remote areas where in-situ observations are sparse \citep{flannigan:86,giglio:03,schroeder:08}. Daily maps of fire radiative power (FRP) \citep{kaufman:03,wooster:05} are available from ECMWF since 2003 through the Global Fire Assimilation System (GFAS) \citep{kaiser:2012,digiuseppe:17,digiuseppe:18}. This dataset has been developed in the framework of the Copernicus Atmosphere Monitoring Services (CAMS) and uses observations from the MODIS sensors on board of Terra and Aqua platforms and assumptions on fire evolution to calculate a continuous record of active fires. The GFAS dataset integrates all available FRP observations available in a day over a regular $0.1\deg$ grid. According to \citet{wooster:05}, this provides an indication of the cumulative dry mass available for burning which can be then put into a relationship with fire emissions. In this paper, the FRP products are only used as an observations of fire events. However, FRP values are ignored and only used to derive a mask of fire occurrence based on a minimum detection criteria: $FRP> 0.5 Wm^{-2}$ \citep{kaiser:2012}. A "hit" is recorded if the fire forecast predicts fire danger above the 95\textsuperscript{th} percentile of its historical values (provided by the ERA5 simulations) when a fire really occurred.

@@

The performance of the the fire forecasting systems to reproduce observed FWI values is assessed using deterministic and probabilistic scores. Both the  synop database and ERA5 are  treated as a proxy for observations in the evaluation. To asses the quality of the computation we use traditional deterministic skill scores such as the  mean bias (MB) and the  mean absolute  error (MAE). For a probabilistic assessment, the continuous ranked probability score is also employed (CRPS; \cite{hersbach2000}). These are defined as:
\begin{align}
MB&=\sum_{p=1}^{cases}\left[F_{n=HRES}-O\right]\\
MAE&=\frac{1}{cases}\sum_{p=1}^{cases}\sqrt{\left[(F_{n=HRES}-O)^2]\right]}\\
CRPS=\frac{1}{cases}\sum_{t=1}^{cases}\int_{-\inf}^{+\inf}\left[F_{n}-O)^2\right]dn
\end{align}
where F is the forecast at time step t of N number of forecasts and O is the observed value. While the MB, MAE and the ACC are applied to a single forecasts, the high resolution forecast,  the CRPS takes into account the whole distribution of possible values predicted by the ensemble. It is the continuous extension of the ranked probability score, where $F_n$ is the cumulative distribution function of the predicted ensemble values. Then, the CRPS compares the cumulative probability distribution of the FWI forecasted by the ensemble forecast system to the observation. In this sense the CRPS is sensitive to the mean forecast biases as well as the spread of the ensemble \citep{hersbach2000}.

@@

In these cases a positively oriented score such as "hit rate" may be more useful. Also forecast quality does not always equals forecast value \citep{richardson:00}. A forecast has high quality if it predicts the observed conditions well according to some objective or subjective criteria. It has value if it helps the user to make a better decision in terms of protective actions \citep{cloke:17}. For example predicting high temperature and low precipitation in desert areas might be accurate but carries low information content and therefore limited value. Following these arguments and to gain an appreciation of the potential value of the forecasting system globally we use as a metric the Probability of Detection (POD), which measures the fraction of the observed events that were correctly forecasted ($POD=hits/(hits+misses)$). Therefore, POD only takes into account observed fires and, unlike other skill scores such as the Brier score, does not suffer from the artificial vanishing due to the high number of correct negative and false alarms (see  \citet{stephenson:08,ferro:11} for a discussion on these problems).

@@

```{r packages, echo = FALSE, message = FALSE, results = 'hide'}
# Install caliver from GitHub
# install.packages("devtools", repos = "http://cran.rstudio.com")
# devtools::install_github("ecmwf/caliver")

# Load relevant packages
library("caliver")
library("raster")
library("ggplot2")
library("verification")
library("rowr")
library("dplyr")
library("rgeos")
library("maptools")
library("sp")
library("RColorBrewer")
library("stringr")
library("colorspace")
```

```{r areas, echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
## Get GFED4 regions and countries of interest
dfgfed4 <- data.frame(ID = 1:14,
                      Region = c("BONA", "TENA", "CEAM", "NHSA", "SHSA", "EURO",
                                 "MIDE", "NHAF", "SHAF", "BOAS", "CEAS", "SEAS",
                                 "EQAS", "AUST"),
                      Zone = c("north", "north", "tropics", "tropics", "south",
                               "north", "north", "tropics", "south", "north",
                               "north", "tropics", "tropics", "south"),
                      stringsAsFactors = FALSE)

BasisRegions <- readRDS("../data/BasisRegions_simplified.rds")

# GFED4 regions of interest
EURO <- BasisRegions[BasisRegions$Region == "EURO",]
TENA <- BasisRegions[BasisRegions$Region == "TENA",]
SHSA <- BasisRegions[BasisRegions$Region == "SHSA",]
# Countries of interest
pt <- raster::getData(name = "GADM", country = "Portugal", level = 1)[-c(2,13),]
chile <- raster::getData(name = "GADM", country = "Chile", level = 0)
cali <- raster::getData(name = "GADM", country = "USA", level = 1)[5, ]

# All the GFED4 regions
BasisRegions_join <- BasisRegions
# add id column for join
BasisRegions_join@data$id <- 1:nrow(BasisRegions_join@data)
regionXFort <- fortify(BasisRegions_join, data = BasisRegions_join@data)
regionXPoly <- merge(regionXFort, BasisRegions_join@data,
                     by.x = "id", by.y = "id")  # join data
# Re-order factors
regionXPoly$Region <- factor(as.character(regionXPoly$Region),
                             levels = dfgfed4$Region)
```

```{r Figure1, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure1}GFED4 regional classification and the 3 countries selected to showcase the fire forecast performances (California, Chile, and Portugal)."}
ggplot(aes(x = long, y = lat), data = map_data("world")) +
  geom_polygon(aes(group = group), fill=NA, colour = "grey65") +
  coord_equal() +  theme_bw() + xlab("Longitude") + ylab("Latitude") +
  geom_polygon(data = regionXPoly,
               aes(x = long, y = lat, group = group, fill = factor(Region)),
               alpha = 0.5) +
  scale_fill_manual(name="GFED4 regions",
                    values = c("darkgrey",
                               RColorBrewer::brewer.pal(12,"Paired"),
                               "lightgrey"),
                    labels = dfgfed4$Region) +
  geom_polygon(data = map_data("world", region = "Portugal"),
               aes(x = long, y = lat, group = group, colour = "Portugal"),
               fill = NA) +
  geom_polygon(data = map_data("world", region = "Chile"),
               aes(x = long, y = lat, group = group, colour = "Chile"),
               fill = NA) +
  geom_polygon(data = map_data(map = "state", region = "California"),
               aes(x = long, y = lat, group = group, colour = "California"),
               fill = NA) +
  scale_colour_manual(name = "Study areas",
                      values = c("Portugal" = "brown",
                                 "California" = "darkblue",
                                 "Chile" = "darkgreen"))
```

@@

\subsection{Skill in the FWI prediction}
The first assessment looks at the capability of ECMWF fire forecast to reproduce the same FWI values as would be estimated from the network of local stations but 2, 6 and 10 days ahead. The selected stations (figure \ref{fig:Figure2}) have at least 30 records during 2017 at local noon and are used to perform an analysis of bias and anomaly correlation at different lead times. For comparison also FWI calculation using ERA5 is included. This provide an estimations of the limit of predictability when using forcing from model simulations in place of observed values \citep{digiuseppe:16}. As expected there is a performance degradation going towards longer lead times and mean biases are limited to few units even at day 10. However, depending on the calibration procedure adopted few units could mean a mismatch in danger level classification.  The mean absolute error  (figure \ref{fig:synop}b)  provides  information on the residual amplitudes.  FWI from reanalysis have the largest skills as expected and the mean absolute errors rapidly increases with lead times. However the distribution of MAE values clearly shows that in selected cases predictive skills can be achievable even at day 10.

```{r Figure2, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure2}Spatial distribution of weather stations from the synop network  which have at least 30 observations recorded at local noon in 2017."}
# Get data from previous publication
# https://github.com/cvitolo/GEFF-ERA5/blob/master/data/df_all.rds
# save them in the "../data" folder
# Use QGIS to beautify the map
knitr::include_graphics("../images/synops_col.png")
```

```{r prepFigure3, echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
## Get data from previous publication
# https://github.com/cvitolo/GEFF-ERA5/blob/master/data/df_geff_erai_era5.rds
# save it in the "../data" folder
# df <- readRDS(file = "../data/df_geff_erai_era5.rds")
# stations <- unique(df[, c("id", "lat", "long")])
# 
# dates2017 <- seq.Date(from = as.Date("2017-01-01"),
#                       to = as.Date("2017-12-31"),
#                       by = "day")

# Get forecasts for day 2-6-10
# This code chunk uses forecast data, which is not open.
# The code is commented out and the result is provided in the form of a table
# and stored in the ../data folder.

# day2FWI <- raster::stack()
# day6FWI <- raster::stack()
# day10FWI <- raster::stack()
# for (file in list.files(path = "/hugetmp/FireEventsDatabase/fwi2017/ffwi_rotated",
#                         pattern = "*hr.nc", full.names = TRUE)){
#   print(file)
#   x <- raster::brick(file)
#   day2FWI <- raster::stack(day2FWI, x[[2]])
#   day6FWI <- raster::stack(day6FWI, x[[6]])
#   day10FWI <- raster::stack(day10FWI, x[[10]])
# }

# df_used <- data.frame(matrix(NA, nrow = 0, ncol = 18))
# names(df_used) <- c("id", "lat", "long", "tzid", "season", "yr", "mon", "day",
#                     "temp", "prec", "rh", "ws", "OBS", "ERAI", "ERA5",
#                     "HRES_d2", "HRES_d6", "HRES_d10")
# for (i in seq_along(stations$id)){
# 
#   # We filter over the station
#   dfx <- df %>% filter(id == stations$id[i],
#                        lat == stations$lat[i],
#                        long == stations$long[i])
#   print(paste(i, dim(dfx)[1]))
# 
#   if (dim(dfx)[1] >= 30){
#     # Calculate the observed FWI
#     initial_condition <- data.frame(ffmc = 85, dmc = 6, dc = 15,
#                                     lat = dfx$lat[1])
#     dfx$fwiobs <- cffdrs::fwi(input = dfx,
#                               init = initial_condition,
#                               out = "fwi")$FWI
# 
#     # Define spatial point
#     spdf <- dfx[1, ]
#     # Promote data.frame to spatial
#     coordinates(spdf) = ~long + lat
# 
#     # Select timestamp
#     mytimestamps <- which(dates2017 %in%
#                             as.Date(paste0(dfx$yr, "-",
#                                            sprintf("%02d", dfx$mon), "-",
#                                            sprintf("%02d", dfx$day))))
# 
#     # Extract the modelled FWI from HRES day 2
#     temp_d2 <- t(raster::extract(x = day2FWI, y = spdf))
#     dfx$HRES_d2 <- temp_d2[mytimestamps]
# 
#     # Extract the modelled FWI from HRES day 6
#     temp_d6 <- t(raster::extract(x = day6FWI, y = spdf))
#     dfx$HRES_d6 <- temp_d6[mytimestamps]
# 
#     # Extract the modelled FWI from HRES day 10
#     temp_d10 <- t(raster::extract(x = day10FWI, y = spdf))
#     dfx$HRES_d10 <- temp_d10[mytimestamps]
# 
#     df_used <- dplyr::bind_rows(df_used, dfx)
#   }
# 
# }
# 
# # Remove NAs
# df_used <- df_used[complete.cases(df_used),]
# saveRDS(df_used, "data/SYNOPtable.rds")

df_used <- readRDS("../data/SYNOPtable.rds")

d <- df_used %>%
  select(fwiobs, ERA5, HRES_d2, HRES_d6, HRES_d10) %>%
  reshape2::melt(id.vars = "fwiobs") %>%
  filter(fwiobs < 250) %>%
  filter(value < 250) %>%
  filter(complete.cases(.)) %>%
  mutate(BIAS = fwiobs - value,
         MAE = abs(fwiobs - value)) %>%
  rename(modelled_type = variable, modelled_value = value) %>%
  reshape2::melt(id.vars = c("fwiobs", "modelled_type", "modelled_value"))
```

```{r Figure3, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure3}Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias (left panel) and the 2017 mean absolute error (right panel)."}
ggplot(d, aes(x = modelled_type, y = value, fill = modelled_type)) +
  facet_wrap( ~ variable, ncol = 2) +
  geom_boxplot(color="black", size=0.2,  outlier.shape = NA) +
  theme_bw() + ylim(-2, 12) +
  xlab("") + ylab("") +
  scale_fill_discrete_qualitative() +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("ERA5", "HRES_D2", "HRES_D6", "HRES_D10"))
```

@@

```{r Figure4, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure4}PUT here crps skills scores."}

```

\subsection{Skill in detecting fire events}

Conventionally, we assume that an active fire is correctly predicted if the FWI is greater than the 95$^{th}$ percentile of its distribution of values here defined using the ERA5 database. Figure~\ref{fig:pod} shows the mean POD for all events in 2017 at forecast day 2, 6 and day 10. Figure~\ref{fig:pod} also shows the POD that could be achieved in the absence of a forecasting system when just using a mean climatological FWI estimation. Given the intrinsic limitations of the POD as skill metric, CLIM provides an useful benchmark to understand the incremental skill provided by the forecast.

At day 2, high latitudes are characterized by POD greater than or equal to 0.5. These are mostly temperate regions where vegetation is dominated by forests and fuel is abundant and where fire danger is moisture limited. In these regions the FWI is a good predictor of fire danger \citep{digiuseppe:16}. It has to be noted that the FWI does not take into account management measures that could introduce a relevant number of "false-alarm". Central America, the Middle East and the northern hemisphere areas, Africa are characterized by a POD in the range 0.3-0.4 as in most of the tropics, where, fires usually occur in grass-shrub lands. Here fuel is scarce and weather plays a less relevant controlling role. Also it has to be noted that  the statistics here are likely to be contaminated by  many agricultural and prescribed fires that are considered “events” and which would dilute some of the skill in regions where annual cropland is high or are heavily managed. The spatial distribution of POD at day 6 is very similar to the corresponding figures at day 2, just slightly shifted towards lower values.

>>>>

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% copernicus_rticles_template (flag for rticles template detection - do not remove!)

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[, manuscript]{copernicus}

%% Journal abbreviations (please use the same for discussion papers and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% ASTRA Proceedings (ap)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

% The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
%

\begin{document}

\title{Fire danger: the skill provided by the ECMWF Integrated Forecasting
System}

\Author[1]{Francesca}{Di Giuseppe}
\Author[1]{Claudia}{Vitolo}
\Author[1]{Blazej}{Krzeminski}
\Author[2]{Jesus}{San-Miguel}
\Author[1]{Florian}{Pappenberger}

\affil[1]{European Centre for Medium-range Weather Forecasts, Reading, United
Kingdom}
\affil[2]{European Commission, Joint Research Centre, Ispra, Italy}

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

\runningtitle{Fire danger forecasting}

\runningauthor{Di Giuseppe et al.}

\correspondence{Francesca\ Di Giuseppe\ (F.DiGiuseppe@ecmwf.int)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
In the framework of the EU Copernicus program the European Centre for
Medium-range Weather Forecast (ECMWF) on behalf of the Joint Research
Centre (JRC) is forecasting daily fire weather indices using its medium
range ensemble prediction system. The use of weather forecast in place
of local observations can extend early warnings up to 1-2 weeks allowing
for greater proactive coordination of resource-sharing and mobilization
within and across countries. In addition, the use of an ensemble system
allows to perform a probabilistic assessment of the forecast
uncertainties which can boost confidence in the decision process during
emergency situations. Using one year of pre-operational service in 2017
here we assess the capability of the system globally and analyze in
detail three major events in Chile, Portugal and California. We also
present examples on how fire forecast products could be tailored to
provide uncertainties of fire weather conditions and information in a
probabilistic fashion.
\end{abstract}

@@

Systems such as the FWI detect dangerous weather conditions conducive of
uncontrollable fires rather than modelling the probability of ignition
and fire behaviours. The FWI (developed in Canada) is specifically
calibrated to describe the fire behaviour in a jack pine stand
(\emph{Pinus banksiana}) typical of the Canadian forests. However, its
simplicity of implementation has made it a popular choice in many
countries and it has shown to perform reasonably well in ecosystems very
dissimilar to the boreal forest \citep{digiuseppe:16,degroot:07}. The
FWI calculation only relies on weather forcings and no information on
the actual vegetation status is taken into account. When weather
forecasts are used in place of observations, uncertainties can be
introduced. Sources of uncertainty can be: (i) the limited
predictability of atmospheric dynamics which is strongly affected by the
initial state and (ii) the misrepresentation of physical processes. In
the former case, errors are randomly distributed around the true state
\citep{orrell:01}; in the latter, errors produce systematic deviations
from the true state. In both cases, errors in the weather forecast may
be amplified or damped by nonlinear transformations in the fire weather
model \citep{erickson:18}. Thus, for example, a dry bias in the model in
a certain region will lead to the persistent prediction of higher fire
danger values compared to what would be calculated using local
observations.

@@

\subsection{FWI calculation}
\subsubsection{General concept}

The Fire Weather Index system is composed of six variables describing
fuel moisture and fire behavior, as influenced by weather
\citep{vanwagner:87}. Three fuel moisture codes provide the moisture
content of dead woody debris of different diameter classes laying on
three fuel beds. The dead fine fuels is described by the Fine Fuel
Moisture Code (FFMC) which has a fast response time to weather
modification (hourly time scales). The surface organic matter of
moderate density, such as the fermentation layer of forest soils
\citep{hood:10} is calculated using the Duff Moisture Code (DMC). DMC
has a slower response to the weather forcing, conventionally set to 12
days. Finally the deep, compact soil organic layers, such as the humus
layer of forest soils \citet{hood:10} is expressed with a Drought Code
(DC). This is the slower evolving component of the FWI codes and has a
lag-time set to 52 days in the original FWI formulation. The FWI system
also provides fire behavior indexes, in terms of rate of fire spread
(Initial Spread Index, ISI), fuel available for combustion (Buildup
Index, BUI), and head fire intensity (Fire Weather Index, FWI). The FWI
component combines the ISI and BUI and can be considered as the most
general indicator of fire danger.

@@

In addition, from ERA5 we derive a climatological benchmark simulation
(called CLIM hereafter), considered as a ``mean year'' for the FWI using
the period 1980-2018. At pixel level, every day the median FWI is
calculated over the period spanning 4 days before/after a given date.
While CLIM has no expected predictive skill for a specific year, it
retains information on the yearly climatological variation of FWI. CLIM
is what could be used in the absence of either a monitoring (based on
observations) or a forecast system (based on weather forecast). It
should therefore score better or equal of the forecast on time ranges
beyond the limits of predictability. CLIM is used in this study as a
benchmark to rank the expected improvements provided by a forecasting
system

@@

While national inventories of wildfire activities exist in many
countries, they can be heterogeneous and lack the temporal span
desirable for the validation of a fire danger system at the global
scale. Satellite observations can supply a valid alternative especially
as they cover remote areas where in-situ observations are sparse
\citep{flannigan:86,giglio:03,schroeder:08}. Daily maps of fire
radiative power (FRP) \citep{kaufman:03,wooster:05} are available from
ECMWF since 2003 through the Global Fire Assimilation System (GFAS)
\citep{kaiser:2012,digiuseppe:17,digiuseppe:18}. This dataset has been
developed in the framework of the Copernicus Atmosphere Monitoring
Services (CAMS) and uses observations from the MODIS sensors on board of
Terra and Aqua platforms and assumptions on fire evolution to calculate
a continuous record of active fires. The GFAS dataset integrates all
available FRP observations available in a day over a regular \(0.1\deg\)
grid. According to \citet{wooster:05}, this provides an indication of
the cumulative dry mass available for burning which can be then put into
a relationship with fire emissions. In this paper, the FRP products are
only used as an observations of fire events. However, FRP values are
ignored and only used to derive a mask of fire occurrence based on a
minimum detection criteria: \(FRP> 0.5 Wm^{-2}\) \citep{kaiser:2012}. A
``hit'' is recorded if the fire forecast predicts fire danger above the
95\textsuperscript{th} percentile of its historical values (provided by
the ERA5 simulations) when a fire really occurred.

@@

The performance of the the fire forecasting systems to reproduce
observed FWI values is assessed using deterministic and probabilistic
scores. Both the synop database and ERA5 are treated as a proxy for
observations in the evaluation. To asses the quality of the computation
we use traditional deterministic skill scores such as the mean bias (MB)
and the mean absolute error (MAE). For a probabilistic assessment, the
continuous ranked probability score is also employed (CRPS;
\cite{hersbach2000}). These are defined as: \begin{align}
MB&=\sum_{p=1}^{cases}\left[F_{n=HRES}-O\right]\\
MAE&=\frac{1}{cases}\sum_{p=1}^{cases}\sqrt{\left[(F_{n=HRES}-O)^2]\right]}\\
CRPS=\frac{1}{cases}\sum_{t=1}^{cases}\int_{-\inf}^{+\inf}\left[F_{n}-O)^2\right]dn
\end{align} where F is the forecast at time step t of N number of
forecasts and O is the observed value. While the MB, MAE and the ACC are
applied to a single forecasts, the high resolution forecast, the CRPS
takes into account the whole distribution of possible values predicted
by the ensemble. It is the continuous extension of the ranked
probability score, where \(F_n\) is the cumulative distribution function
of the predicted ensemble values. Then, the CRPS compares the cumulative
probability distribution of the FWI forecasted by the ensemble forecast
system to the observation. In this sense the CRPS is sensitive to the
mean forecast biases as well as the spread of the ensemble
\citep{hersbach2000}.

@@

In these cases a positively oriented score such as ``hit rate'' may be
more useful. Also forecast quality does not always equals forecast value
\citep{richardson:00}. A forecast has high quality if it predicts the
observed conditions well according to some objective or subjective
criteria. It has value if it helps the user to make a better decision in
terms of protective actions \citep{cloke:17}. For example predicting
high temperature and low precipitation in desert areas might be accurate
but carries low information content and therefore limited value.
Following these arguments and to gain an appreciation of the potential
value of the forecasting system globally we use as a metric the
Probability of Detection (POD), which measures the fraction of the
observed events that were correctly forecasted
(\(POD=hits/(hits+misses)\)). Therefore, POD only takes into account
observed fires and, unlike other skill scores such as the Brier score,
does not suffer from the artificial vanishing due to the high number of
correct negative and false alarms (see \citet{stephenson:08,ferro:11}
for a discussion on these problems).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{article_files/figure-latex/Figure1-1} \caption{\label{fig:Figure1}GFED4 regional classification and the 3 countries selected to showcase the fire forecast performances (California, Chile, and Portugal).}\label{fig:Figure1}
\end{figure}

@@

\subsection{Skill in the FWI prediction}

The first assessment looks at the capability of ECMWF fire forecast to
reproduce the same FWI values as would be estimated from the network of
local stations but 2, 6 and 10 days ahead. The selected stations (figure
\ref{fig:Figure2}) have at least 30 records during 2017 at local noon
and are used to perform an analysis of bias and anomaly correlation at
different lead times. For comparison also FWI calculation using ERA5 is
included. This provide an estimations of the limit of predictability
when using forcing from model simulations in place of observed values
\citep{digiuseppe:16}. As expected there is a performance degradation
going towards longer lead times and mean biases are limited to few units
even at day 10. However, depending on the calibration procedure adopted
few units could mean a mismatch in danger level classification. The mean
absolute error (figure \ref{fig:synop}b) provides information on the
residual amplitudes. FWI from reanalysis have the largest skills as
expected and the mean absolute errors rapidly increases with lead times.
However the distribution of MAE values clearly shows that in selected
cases predictive skills can be achievable even at day 10.

\begin{figure}
\includegraphics[width=1\linewidth]{../images/synops_col} \caption{\label{fig:Figure2}Spatial distribution of weather stations from the synop network  which have at least 30 observations recorded at local noon in 2017.}\label{fig:Figure2}
\end{figure}

\begin{verbatim}
## Warning: Removed 958035 rows containing non-finite values (stat_boxplot).
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{article_files/figure-latex/Figure3-1} \caption{\label{fig:Figure3}Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias (left panel) and the 2017 mean absolute error (right panel).}\label{fig:Figure3}
\end{figure}

@@

\subsection{ Skill in detecting fire events}

Conventionally, we assume that an active fire is correctly predicted if
the FWI is greater than the 95\(^{th}\) percentile of its distribution
of values here defined using the ERA5 database.
Figure\textasciitilde{}\ref{fig:pod} shows the mean POD for all events
in 2017 at forecast day 2, 6 and day 10.
Figure\textasciitilde{}\ref{fig:pod} also shows the POD that could be
achieved in the absence of a forecasting system when just using a mean
climatological FWI estimation. Given the intrinsic limitations of the
POD as skill metric, CLIM provides an useful benchmark to understand the
incremental skill provided by the forecast.

At day 2, high latitudes are characterized by POD greater than or equal
to 0.5. These are mostly temperate regions where vegetation is dominated
by forests and fuel is abundant and where fire danger is moisture
limited. In these regions the FWI is a good predictor of fire danger
\citep{digiuseppe:16}. It has to be noted that the FWI does not take
into account management measures that could introduce a relevant number
of ``false-alarm''. Central America, the Middle East and the northern
hemisphere areas, Africa are characterized by a POD in the range 0.3-0.4
as in most of the tropics, where, fires usually occur in grass-shrub
lands. Here fuel is scarce and weather plays a less relevant controlling
role. Also it has to be noted that the statistics here are likely to be
contaminated by many agricultural and prescribed fires that are
considered ``events'' and which would dilute some of the skill in
regions where annual cropland is high or are heavily managed. The
spatial distribution of POD at day 6 is very similar to the
corresponding figures at day 2, just slightly shifted towards lower
values.

One important exception is the extremely good performance of the fire
forecast in Equatorial Asia where the system seems to have a
predictability of 0.9 even at day 6. Also \citet{degroot:07} highlighted
how FWI is not a good indicator in this area and a fire early warning
system should mostly rely on the drought code. Still, there are factors
that could contribute to this enhanced predictability. First fires in
this region are mainly caused by humans for the purposes of cleaning the
land for establishing plantations \citep{field:09,benedetti:16}.
Therefore they occur every season as soon as fuel conditions are
favorable and the ignition is less of a stochastic component. This also
means that the POD statistics are built on a much higher number of
events. Moreover, the strength and prevalence of these fires are
strongly influenced by large-scale climate patterns like El
Ni\{\textasciitilde{}n\}o \cite{field:04} which have been proven being
highly predictable \citep{zhu:15}. The other relevant fact is the almost
null skill globally when a climatological FWI is used highlighting the
added benefit of a forecasting system
